---
title: "King County Housing Prices"
author: "Duncan McKinnon"
output: html_notebook
---

This project aims to explore data on housing prices in an effort to design a model for estimating pricing and other behaviors of the housing market, specifically within King County. 


```{r message = F, warning = F, echo = F, strip.white = F, tidy = T}

suppressPackageStartupMessages
(
  {
    #install or load required packages into the workspace
    package_list <- c('tidyverse', 'knitr', 'plotly','RColorBrewer','sqldf')
    non_installed <- package_list[!(package_list %in% installed.packages()[,"Package"])]
    if(length(non_installed)) install.packages(non_installed)
    #library('tidyverse')
    library('plotly')
    library('RColorBrewer')
    library('sqldf')
    library('knitr')
  }
)
knitr::opts_chunk$set(message = F, warning = F, strip.white = F, tidy = T)
```
load in the required packages


```{r}
#load dataset from project folder
suppressWarnings({housing_data <- read.csv("kc_house_data.csv", stringsAsFactors = F)})


#remove unused columns and normalize the data
housing_data_rev <- housing_data[, 3:21]
```
load in dataset and remove unused features


```{r}
#collect a random sampling of the dataset ordered by price
train_vals <- sample(dim(housing_data)[1], 1000)
test_vals <- sample(dim(housing_data)[1], 200)

#full test set?
#test_vals <- !(1:dim(housing_data)[1] %in% train_vals)

housing_train_data <- as.data.frame(housing_data_rev[train_vals,])
housing_train_data <- housing_train_data[order(-housing_train_data$price),]
housing_test_data <- as.data.frame(housing_data_rev[test_vals,])
housing_test_data <- housing_test_data[order(-housing_test_data$price),]

freq_train <- hist(housing_train_data$price, 
             n = 1000, plot = F)

sqldf("Select price As Pricing, grade As Rating, condition As Condition, sqft_lot As Sq_Footage 
       From housing_train_data
       Limit 20")

```
This dataset is very large, so in order to get a better visual representation of the distribution of the data, a sample of 1000 randomly chosen rows is used instead of the whole dataset (ordered by price, as this is the dependent variable).  The sample indexes are selected at runtime every time the script runs, so the data and analysis are different each time the script is run. Because the sampling is random and the dataset is homogenous and static, the 1000 rows chosen should always be representative of the nature of the whole dataset.


```{r}
#Look at the distributions:
p1 <- plot_ly() %>% add_markers(name = "Pricing",
                                x = 1:length(housing_train_data$price), 
                                y = housing_train_data$price) %>% layout(title = "King County Housing Feature Distributions")
p2 <- plot_ly() %>% add_markers(name = "Rating v Pricing", 
                                x = housing_train_data$grade, 
                                y = housing_train_data$price)
p3 <- plot_ly() %>% add_markers(name = "Condition v Pricing", 
                                x = housing_train_data$condition, 
                                y = housing_train_data$price)
p4 <- plot_ly() %>% add_markers(name = "Sq. Footage v Pricing", 
                                x = housing_train_data$sqft_lot, 
                                y = housing_train_data$price)

subplot(p1, p2, p3, p4, nrows = 2)
```
Looking at the distributions of prices over their ordering by magnitude, it appears that the prices of houses that are below \$1-2 million follow a linear decline, with all houses above \$1-2 million increasing sharply in their prices and bucking the trend.  Assuming that the features in this dataset capture most of the factors that influence the pricing of a home, some linear combination of the features could give a reasonable estimate for the pricing of an individual house below \$1-2 million.  Since the distribution is closer to a power law, two seperate linear functions should give a pretty good estimate by accounting for the 'knee' in the pricing distribution.  First though, we'll try fitting a single line to the data and see what kind of error comes out.

```{r message = F, warning = F}
#split data for training
#test general linear model on relevant quantitative data

housing_train_price <- housing_train_data[, "price"]
housing_test_price <- housing_test_data[, "price"]


glm_mod <- glm(price ~ sqft_living + sqft_lot + condition + grade + yr_built + yr_renovated, data = housing_train_data)

eqn <- paste("Price = Living Space * ", glm_mod$coefficients[2],
             " + Lot Space * ",glm_mod$coefficients[3],
             " + Condition * ", glm_mod$coefficients[4],
             " + Rating * ", glm_mod$coefficients[5],
             " + Year Built * ", glm_mod$coefficients[6],
             " + Year Renovated * ", glm_mod$coefficients[7],
             " + ", glm_mod$coefficients[1])

price_mod <- predict(glm_mod, newdata = housing_test_data)
xvals <- 1:length(price_mod)

price_est_plotly <- plot_ly() %>% 
  add_markers(name = "Test Set Price", x = xvals, y = housing_test_price) %>%
  add_markers(name = "Linear Model est", x = xvals, y = price_mod) %>%
  add_lines(name = "Magnitude Difference", x = xvals, y = price_mod - housing_test_price) %>%
  layout(title = "Attempting Linear Regression Model
         for Pricing Homes in King County")

price_est_plotly
#knit_print(eqn)
```
Given that the distribution of prices was already known to follow a non-linear distribution, the result of this linear estimate isn't very surprising.  The linear fit may have been within a few hundred thousand dollars for the long tail, but the whole seems to be corrupted in trying to fit the collection of values above the elbow.  Below around \$1 million dollars, the housing prices are pretty evenly distributed, but as the prices increase they start jumping up by orders of magnitude.  Lets try the same approach for all values of price < $1 million

```{r}
housing_train_data_rev1 <- housing_train_data[housing_train_data$price < 1e6, ]
housing_train_price_rev1 <- housing_train_price[housing_train_price < 1e6]
housing_test_data_rev1 <- housing_test_data[housing_test_data$price < 1e6, ]
housing_test_price_rev1 <- housing_test_price[housing_test_price < 1e6]

glm_mod_rev1 <- glm(price ~ sqft_living + sqft_lot + condition + grade + yr_built + yr_renovated, data = housing_train_data_rev1)

eqn_rev1 <- paste("Price = Living Space * ", glm_mod_rev1$coefficients[2],
             " + Lot Space * ",glm_mod_rev1$coefficients[3],
             " + Condition * ", glm_mod_rev1$coefficients[4],
             " + Rating * ", glm_mod_rev1$coefficients[5],
             " + Year Built * ", glm_mod_rev1$coefficients[6],
             " + Year Renovated * ", glm_mod_rev1$coefficients[7],
             " + ", glm_mod_rev1$coefficients[1])

price_mod_rev1 <- predict(glm_mod_rev1, newdata = housing_test_data_rev1)
xvals_rev1 <- 1:length(price_mod_rev1)

price_est_plotly_rev1 <- plot_ly() %>% 
  add_markers(name = "Test Set Price", x = xvals_rev1, y = housing_test_price_rev1) %>%
  add_markers(name = "Linear Model est", x = xvals_rev1, y = price_mod_rev1) %>%
  add_lines(name = "Magnitude Difference", x = xvals_rev1, y = price_mod_rev1 - housing_test_price_rev1,
            text = eqn) %>%
  layout(title = "Attempting Linear Regression Model 
         for Pricing Homes in King County
         Where Price of Home < $1 million")

price_est_plotly_rev1
#knit_print(eqn_rev1)
```
```{r}

```















