---
title: "King County Housing Prices"
author: "Duncan McKinnon"
output: html_notebook
---

This project aims to explore data on housing prices in an effort to design a model for estimating pricing and other behaviors of the housing market, specifically within King County. 


```{r}

suppressPackageStartupMessages
(
  {
    #install or load required packages into the workspace
    package_list <- c('tidyverse', 'knitr', 'plotly','RColorBrewer','sqldf')
    non_installed <- package_list[!(package_list %in% installed.packages()[,"Package"])]
    if(length(non_installed)) install.packages(non_installed)
    #library('tidyverse')
    library('plotly')
    library('RColorBrewer')
    library('sqldf')
    library('knitr')
  }
)
knitr::opts_chunk$set(message = F, warning = F, strip.white = F, tidy = T)
```
load in the required packages


```{r}
#load dataset from project folder
suppressWarnings({housing_data <- read.csv("kc_house_data.csv", stringsAsFactors = F)})


#remove unused columns and normalize the data
housing_data_rev <- housing_data[, 3:21]
```
load in dataset and remove unused features


```{r}
#collect a random sampling of the dataset ordered by price
sample_values <- sample(dim(housing_data)[1], 200)

housing_data_sample <- as.data.frame(housing_data_rev[sample_values,])

housing_data_sample <- sqldf("Select * 
                           From housing_data_sample
                           Order By price desc")

sqldf("Select price As Pricing, grade As Rating, sqft_lot As Sq_Footage, yr_built As Year_Built
       From housing_data_sample
       Limit 20")

```
This dataset is very large, so in order to get a better visual representation of the distribution of the data, a sample of 200 randomly chosen rows is used instead of the whole dataset (ordered by price, as this is the dependent variable).  The sample indexes are selected at runtime every time the script runs, so the data and analysis are different each time the script is run. Because the sampling is random and the dataset is homogenous and static, the 200 rows chosen should always be representative of the nature of the whole dataset.


```{r}
#Look at the distributions:
p1 <- plot_ly() %>% add_markers(name = "Pricing", x = 1:200, y = housing_data_sample$price) %>% layout(title = "King County Housing Feature Distributions")
p2 <- plot_ly() %>% add_markers(name = "Rating", x = housing_data_sample$price, y = housing_data_sample$grade)
p3 <- plot_ly() %>% add_markers(name = "Sq. Footage", x = housing_data_sample$price, y = housing_data_sample$sqft_lot)
p4 <- plot_ly() %>% add_markers(name = "Year Built", x = housing_data_sample$price, y = housing_data_sample$yr_built)
subplot(p1, p2, p3, p4, nrows = 2)
```



```{r message = F, warning = F}
#split data for training
#test general linear model on relevant quantitative data
train_set_glm <- housing_data_sample[, !(colnames(housing_data_sample) %in% c("price", "yr_renovated", "lat", "long", "zipcode"))]
train_price <- housing_data_sample[, "price"]

glm_mod <- glm.fit(x = train_set_glm, y = train_price)

price_mod <- fitted(glm_mod)

price_est_plotly <- plot_ly() %>% 
  add_markers(name = "actual price", x = 1:length(train_price), y = train_price) %>%
  add_markers(name = "linear model est", x = 1:length(price_mod), y = price_mod) %>%
  add_lines(name = "magnitude difference", x = 1:length(train_price), y = price_mod - train_price) %>%
  layout(title = "Attempted Linear Regression Model </br> 
         for Pricing Homes in King County </br>")

price_est_plotly
```

