
---
title: "How Frequent Is My Myers-Briggs Type?"
author: "Duncan McKinnon (INFJ/INTJ)"
output: html_notebook
---

This notebook uses a dataset on the frequency of each Myers-Briggs personality type to look at how each of the traits that come together to identify a set of personality types are distributed in the population.

```{r list tables}
library(data.world)
library(stringr)

knitr::opts_chunk$set(message = F, warning = F, strip.white = F, tidy = T)

dataset_key <- "https://data.world/makeovermonday/how-frequent-is-my-myers-briggs-type"
mb_qry <- data.world::qry_sql("SELECT * FROM myersbriggstypes")
mb_df <- data.world::query(mb_qry, dataset = dataset_key)

mb_df$t_hinking_f_eeling <- ifelse(mb_df$t_hinking_f_eeling == "true", "T", "F")

# Add full type names
types <- c()
for(i in 1:16)
{
  types <- c(types, str_c(mb_df[i,1], mb_df[i,2], mb_df[i,3], mb_df[i,4]))
}

mb_df$meyers_briggs <- types

mb_df <- data.frame(mb_df[,1], mb_df[,2], mb_df[,3], mb_df[,4], mb_df[,6], mb_df[,5])

mb_df
```

This pie chart gives a proportioned representation of the number of people who fit into each of the 16 types after taking the meyers briggs test.
```{r pie_chart_overall}
library(plotly)

pie_overall <- plot_ly(mb_df) %>% add_pie(values = ~ratio_in_population, labels = ~meyers_briggs) %>% 
                               layout(title = "Myers-Briggs Personality Type Breakdown", showlegend = F)



pie_overall

```
We can break this down further into each of the individual traits that make up the myers briggs type to look at how individual's scores are distributed here.  The pie charts below show how each trait (EI/SN/TF/IJ) is is divided within the population.
```{r pie_charts_by_trait}
library(sqldf)

# Function to generate a data frame of this trait's ratio and labels
trait_ratio <- function(col, type, labels)
{
  val_A <- unlist(sqldf(str_c("SELECT SUM(ratio_in_population) FROM mb_df
                           WHERE ", col , " == \'", type, "\'")))
  val_B <- 1 - val_A
  
  return(data.frame('labels' = labels, 'values' = c(val_A[1], val_B[1])) )
}

# List of trait data frames
mb_traits <- list()

mb_traits[["E/I"]] <- trait_ratio("e_xtroversion_i_ntroversion", "E", c("Extroversion", "Introversion"))
mb_traits[["S/N"]] <- trait_ratio("s_ensing_i_n_tuition", "S", c("Sensing", "Intuition"))
mb_traits[["T/F"]] <- trait_ratio("t_hinking_f_eeling", "T", c("Thinking", "Feeling"))
mb_traits[["J/P"]] <- trait_ratio("j_udging_p_erceiving", "J", c("Judging", "Perceiving"))


# List of trait pie-charts
plot_list <- plot_ly() %>% add_pie(data = mb_traits[["E/I"]], values = ~values, labels = ~labels, name = "Extroversion / Introversion",
                                   domain = list(x = c(0, 0.4), y = c(0.6, 1))) %>%
                           add_pie(data = mb_traits[["S/N"]], values = ~values, labels = ~labels, name = "Sensing / Intuition",
                                   domain = list(x = c(0.6, 1), y = c(0.6, 1))) %>%
                           add_pie(data = mb_traits[["T/F"]], values = ~values, labels = ~labels, name = "Thinking / Feeling",
                                   domain = list(x = c(0, 0.4), y = c(0, 0.4))) %>%
                           add_pie(data = mb_traits[["J/P"]], values = ~values, labels = ~labels, name = "Judging / Perceiving",
                                   domain = list(x = c(0.6, 1), y = c(0, 0.4))) %>%
                           layout(title = "Myers Briggs By Trait", showlegend = T,
                                  xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
                                  yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))

plot_list
``` 
We can go even deeper into how the traits are distributed by looking at individual combinations of traits.  These are represented by all 2 placed letter combinations (traits) within a myers briggs type, with 4 possible combinations for each pair of traits.  The charts below show how each pair of traits is divided within the population.    
```{r pie_charts_combined_traits}

mb_trait_comb <- list()

# Function to generate a data frame of this trait's ratio and labels
trait_ratio_comb <- function(col, type, labels)
{
  
  str_parts <- c()
  str_parts <- c(str_parts, str_c(col[1] , " == \'", type[1], "\' AND ", col[2], " == \'", type[3], "\'"))
  str_parts <- c(str_parts, str_c(col[1] , " == \'", type[1], "\' AND ", col[2], " == \'", type[4], "\'"))
  str_parts <- c(str_parts, str_c(col[1] , " == \'", type[2], "\' AND ", col[2], " == \'", type[3], "\'"))
  str_parts <- c(str_parts, str_c(col[1] , " == \'", type[2], "\' AND ", col[2], " == \'", type[4], "\'"))
  
  str_tot <- "SELECT SUM(ratio_in_population) FROM mb_df WHERE "
  values = c()
  
  for(i in 1:4)
  {
    qry <- str_c(str_tot, str_parts[i], collapse = "") 
    
    values <- c(values, unlist(sqldf(qry)))
  }
  
  return(data.frame('labels' = labels, 'values' = values ))
}

mb_trait_comb[["EI/TF"]] <- trait_ratio_comb(c("e_xtroversion_i_ntroversion", "t_hinking_f_eeling"), c("E","I","T","F"),
                                             c("Extroversion/Thinking", "Extroversion/Feeling", "Introversion/Thinking", "Introversion/Feeling"))

mb_trait_comb[["SN/JP"]] <- trait_ratio_comb(c("s_ensing_i_n_tuition", "j_udging_p_erceiving"), c("S", "N", "J", "P"),
                                             c("Sensing/Judging", "Sensing/Perceiving", "Intuition/Judging", "Intuition/Perceiving"))

mb_trait_comb[["EI/SN"]] <- trait_ratio_comb(c("e_xtroversion_i_ntroversion", "s_ensing_i_n_tuition"), c("E","I","S","N"),
                                             c("Extroversion/Sensing", "Extroversion/Intuition", "Introversion/Sensing", "Introversion/Intuition"))

mb_trait_comb[["TF/JP"]] <- trait_ratio_comb(c("t_hinking_f_eeling", "j_udging_p_erceiving"), c("T", "F", "J", "P"),
                                             c("Thinking/Judging", "Thinking/Perceiving", "Feeling/Judging", "Feeling/Perceiving"))

mb_trait_comb[["JP/EI"]] <- trait_ratio_comb(c("j_udging_p_erceiving", "e_xtroversion_i_ntroversion"), c("J", "P", "E", "I"),
                                             c("Judging/Extroversion", "Judging/Introversion", "Perceiving/Extroversion", "Perceiving/Introversion"))

mb_trait_comb[["SN/TF"]] <- trait_ratio_comb(c("s_ensing_i_n_tuition", "t_hinking_f_eeling"), c("S", "N", "T", "F"),
                                             c("Sensing/Thinking", "Sensing/Feeling", "Intuition/Thinking", "Intuition/Feeling"))

plot_trait_comb <- plot_ly() %>% add_pie(data = mb_trait_comb[["EI/TF"]], values = ~values, labels = ~labels, name = "EI/TF",
                                    domain = list(x = c(0.1, 0.5), y = c(0.7, 1))) %>%
                                add_pie(data = mb_trait_comb[["SN/JP"]], values = ~values, labels = ~labels, name = "SN/JP",
                                    domain = list(x = c(0.5, 0.9), y = c(0.7, 1))) %>%
                                add_pie(data = mb_trait_comb[["EI/SN"]], values = ~values, labels = ~labels, name = "EI/SN",
                                    domain = list(x = c(0, 0.4), y = c(0.35, 0.65))) %>%
                                add_pie(data = mb_trait_comb[["TF/JP"]], values = ~values, labels = ~labels, name = "TF/JP",
                                    domain = list(x = c(0.6, 1), y = c(0.35, 0.65))) %>%
                                add_pie(data = mb_trait_comb[["JP/EI"]], values = ~values, labels = ~labels, name = "JP/EI",
                                    domain = list(x = c(0.1, 0.5), y = c(0, 0.3))) %>%
                                add_pie(data = mb_trait_comb[["SN/TF"]], values = ~values, labels = ~labels, name = "SN/TF",
                                    domain = list(x = c(0.5, 0.9), y = c(0, 0.3))) %>%
                                layout(title = "Combined Traits Distributions", showlegend = F,
                                    xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
                                    yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))

plot_trait_comb

```

