---
title: "Fatal Police Shootings, 2015-Present"
author: "Duncan McKinnon"
output: html_document
---
#Civilians shot and killed by on-duty police officers in United States
#From Location: https://www.kaggle.com/washingtonpost/police-shootings
#May require membership in data.world to run properly


```{r}
knitr::opts_chunk$set(message = F, warning = F, strip.white = F, tidy = T)
suppressPackageStartupMessages(
  {
    package_list <- c('data.world','plotly', 'maps', 'foreach','RColorBrewer', 'dplyr', 'stringr', 'sqldf', 'magrittr','webshot')
    non_installed <- package_list[!(package_list %in% installed.packages()[,"Package"])]
    if(length(non_installed)) 
    {
      install.packages(non_installed)
      if(is.null(getOption("dw.auth_token")))
      {
        data.world::set_config(data.world::save_config(auth_token = dw_token)) #plug in your token if needed
      }
    }
    library('data.world')
    library('plotly')
    library('maps')
    library('foreach')
    library('RColorBrewer')
    library('tidyverse')
    library('sqldf')
    library('webshot')
  }
)

```
load required R packages

```{r}

#load from data.world
data_key <- "https://data.world/data-society/fatal-police-shootings"
raw_data_qry <- qry_sql("SELECT * FROM Tables")
raw_data_tables <- query(qry = raw_data_qry, data_key)
raw_data <- query(qry_sql(sprintf("Select * From %s", raw_data_tables$tableName[[1]])), data_key)
raw_data <- raw_data[complete.cases(raw_data),]

#Data Cleaning
colnames(raw_data) <- c("ID", "Name", "Date", "Manner_of_Death",
                                   "Armed", "Age", "Gender", "Race",
                                   "City", "State", "Mental_Illness",
                                   "Attack", "Fleeing", "Body_Cam")


raw_data$Race[raw_data$Race == ""] <- "UNK" 

raw_data_clean <- data.frame(raw_data)
raw_data_clean$Date <- as.Date(raw_data$Date)
raw_data_clean$Mental_Illness <- ifelse(raw_data$Mental_Illness == "True", T, F)
raw_data_clean$Attack <- ifelse(raw_data$Attack == "attack", T, F)
raw_data_clean$Armed <- as.logical(ifelse(raw_data$Armed %in% c("unarmed","undetermined"), F, T))
raw_data_clean$Fleeing <- as.factor(raw_data$Fleeing)


raw_data_nonViolent <- sqldf("SELECT * 
                       FROM raw_data_clean
                       WHERE Attack")

raw_data_plotly <- plot_ly(raw_data_nonViolent) %>%
                   add_markers(x = ~Date, y = ~Age, color = ~Race, 
                               colors = "Spectral",
                               text = ~paste("</br>Name: ", Name,
                                             "</br>Race: ", Race,
                                             "</br>Gender: ", Gender)) %>%
                   layout(title = "Date and Age of Victims Killed By Policing in Nonviolent Encounters")

suppressWarnings({raw_data_plotly})

```
This graph represents the date, age and ethnicity of each person shot in the US by police between 2015-2017 and who was also reported as nonviolent.  Although the variables seem distributed accross the features, there are clearly more people below 50 who are shot by killed during this time, and most of them were identified as white.  This fits the trends within the population at a high level, because whites do make up the true majority in the USA.
```{r}
cityShootings <- sqldf("Select COUNT(City) As Shootings, LOWER(City) As City, 
                        State, Race
                         From raw_data_nonViolent
                         Group By City
                         Order By State")

#clean up city names for data join on city and state
cityLoc <- us.cities
cityLoc$name <- foreach(x = 1:length(cityLoc$name), .combine = c) %do%      
                        tolower(rawToChar(charToRaw(cityLoc$name[x])[1:(length(charToRaw(cityLoc$name[x]))-3)]))

colnames(cityLoc) <- c("City", "State", "Population", "Latitude", "Longitude", "Capital")

cityShootingLoc <- sqldf("Select cityShootings.*, Population/cityShootings.Shootings As Killing_Per, 
                          Population, Latitude, Longitude
                          From cityShootings Inner Join cityLoc 
                          On cityLoc.City == cityShootings.City And cityLoc.State == cityShootings.State
                          Order By cityShootings.State, cityShootings.City")




cityShooting_geo_plotly <- ggplot2::map_data("state") %>%
                           plot_ly(x = ~long, y = ~lat)%>%
                           group_by(group) %>%
                           add_polygons(name = "USA", hoverinfo = "none") %>%
                           add_markers(name = "Shootings", data = cityShootingLoc[ifelse(
                             cityShootingLoc$State %in% c('HI','AK'), FALSE, TRUE),], 
                                       x = ~Longitude, y = ~Latitude, 
                                       color = ~Killing_Per, colors = "Spectral",
                                       size = ~Shootings,
                                       text = ~paste("</br>City: ", City,
                                                     "</br>State: ", State,
                                                     "</br>Population: ", Population,
                                                     "</br>Shootings: ", Shootings),
                                       hoverinfo = text) %>%
                          layout(title = "Police Shootings of Non-Violent Civilians 2015 - 2017", 
                                 xaxis = list(title = "Longitude"),
                                 yaxis = list(title = "Latitude"))


cityShooting_geo_plotly
#cityShooting_geo_gg
```

There is no question that police shootings of non-violent civilians are rare events.  This map shows the number of civilians killed in major cities across the US. According to this combined dataset, the chances of any civilian, with any features being killed in any non-violent encounter with police is around 1/100141.  By looking more closely at the features used to identify individuals, we can see how that average holds up across all races, ages and genders, or if people with certain features in certain places are more likely to be killed by police on aggregate.
```{r}
ByRaceCount <- sqldf("SELECT race As Race, COUNT( race ) As Number
                      FROM raw_data_nonViolent
                      GROUP BY race")

#data on race in the US: https://www.census.gov/quickfacts/fact/table/US/PST045216
ByRaceCount$RacePCT_2016 <-  c( "A" = 0.057, 
                                "B" = 0.133, 
                                "H" = 0.178, 
                                "N" = 0.013, 
                                "O" = 0.002,
                                "UNK" = 0.004,
                                "W" = 0.613)

ByRaceCount$Expected_Number <- ceiling(sum(ByRaceCount$Number) * ByRaceCount$RacePCT_2016)

ByRaceCount_plotly <- plot_ly() %>% 
                      add_pie(data = ByRaceCount[c("Race", "Number")], 
                              name = "Actual Result", labels =~Race, values = ~Number,
                              domain = list(x = c(0, 0.4), y = c(0.3, 0.7))) %>% 
                      add_pie(data = ByRaceCount[c("Race", "Expected_Number")],
                              name = "Expected Result", labels = ~Race, values = ~Expected_Number,
                              domain = list(x = c(0.6, 1), y = c(0.3, 0.7))) %>%
                      layout(title = "Police Shootings of Non-Violent Civilians By Race 2015-2017 
                             </br> Actual vs. By Representation in Population
                             </br>(Using 2016 Population Percentages)",
                             xaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE),
                             yaxis = list(showgrid = FALSE, zeroline = FALSE, showticklabels = FALSE))

ByRaceCount_plotly
```
These pie charts show how the total number of non-violent offenders killed by police is distributed by ethnicity compared with the distribution of ethnicities across the population.  If there were no racial bias in the victims of police killings, the percentages of people of each ethnicity killed in non-violent encounters with police would roughly match the racial breakdown of the population in which they all exist because it would be randomized. Instead, this data shows that there is a strong bias for black victims being killed in non-violent encounters, with close to double the expected instances occuring.  The reason for this bias cannot be derived from the data with only the analysis provided thus far, but further exploration will attempt to determine if there is a reasonable cause within the available data.
```{r}
#instances of violent encounters



```







